---
title: "QuickBLAST"
author: "Vishvesh Karthik"
date: "`r Sys.Date()`"
urlcolor: blue
output:
  rmarkdown::html_vignette:
    highlight: espresso
    self_contained: true
    wrap: 60
    columns: 60
    toc: true
    toc_depth: 4
    number_sections: true
vignette: >
  %\VignetteIndexEntry{QuickBLAST}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

---
title: "Introduction to QuickBLAST"
---

R is widely used for data analysis, but running NCBI's standard BLAST tools within R has traditionally been slow. Because the NCBI C++ toolkit is massive and difficult to compile for R, existing packages are forced to run BLAST as an external subprocess, creating major read/write bottlenecks.

QuickBLAST solves this by building a direct bridge between R and the NCBI C++ toolkit via Rcpp. By bypassing traditional text-based formatting and transporting data directly into memory using Apache Arrow, QuickBLAST performs sequence comparisons exceptionally fast.
1. Setup and Instance Management

QuickBLAST operates using "instances". An instance is an active C++ object pointer (QuickBLAST_XPtr) that holds your BLAST configuration open in memory.

```{r setup, echo=TRUE,include=T, tidy=TRUE}
library(QuickBLAST)
#> Loading required package: Rcpp
#> Loading required package: RcppProgress
#> Loading required package: dplyr
#> Loading required package: BH
#> QuickBLAST Loaded!
#> Version: 1.4.7
#> Github: [https://github.com/vizkidd/QuickBLAST](https://github.com/vizkidd/QuickBLAST)

QuickBLAST::isQuickBLASTLoaded()
#> [1] TRUE

# Create instances for different sequence types and programs.
# Note: seq_type = 1 (Protein), seq_type = 0 (Nucleotide)
blastp_inst <- QuickBLAST::CreateQuickBLASTInstance(seq_type = 1, strand = 0, program = "blastp", save_sequences = TRUE, options = "-evalue 100000")
#> Option: evalue set to : 100000

blastn_inst <- QuickBLAST::CreateQuickBLASTInstance(seq_type = 0, strand = 0, program = "blastn", options = "-evalue 100000")
#> Option: evalue set to : 100000

tblastx_inst <- QuickBLAST::CreateQuickBLASTInstance(seq_type = 0, strand = 0, program = "tblastx")
#> Using tblastx Defaults...

# Inspecting an instance reveals its underlying C++ pointer and attributes
blastn_inst
#> <pointer: 0x61241543c840>
#> attr(,"seq_type")
#> [1] 0
#> attr(,"strand")
#> [1] 0
#> attr(,"program")
#> [1] "blastn"
#> attr(,"options")
#> [1] "-evalue 100000"
#> attr(,"save_sequences")
#> [1] FALSE
#> attr(,"save_hsp_sequences")
#> [1] FALSE
#> attr(,"class")
#> [1] "QuickBLAST_XPtr"
```

### Managing Memory

Because QuickBLAST creates objects in C++, we provide convenience functions to track, verify, and clean up instances to prevent memory leaks.

```{r inst_man, echo=TRUE,include=T, tidy=TRUE}
QuickBLAST::GetInstanceCount()
#> [1] 3

QuickBLAST::GetInstanceID(blastp_inst)
#> [1] 0

# Clean up memory by deleting specific instances
QuickBLAST::DeleteQuickBLASTInstance(QuickBLAST::GetQuickBLASTInstance(1)) 
#> [1] TRUE

QuickBLAST::DeleteQuickBLASTInstance(2) 
#> [1] TRUE

QuickBLAST::GetInstanceCount()
#> [1] 1

# Attempting to fetch a deleted instance securely throws a C++ map error
try(identical(QuickBLAST::GetQuickBLASTInstance(1), blastn_inst))
#> Error fetching QuickBLAST instance (1): map::at
#> [1] FALSE

# Recreate the blastn instance for later steps
blastn_inst <- QuickBLAST::CreateQuickBLASTInstance(seq_type = 0, strand = 0, program = "blastn")
#> Using blastn Defaults...
```

## 2) Raw Sequence Comparison

You can pass raw character strings directly to QuickBLAST. The engine evaluates the sequences in memory instantly. Notice that QuickBLAST includes automatic recovery: if a pointer is dead, it will attempt to reload it.

```{r}
# (~)Successful TBLASTX execution : Garbage low information nucleotide sequences
QuickBLAST::BLAST2Seqs(tblastx_inst, "AAAAAAAAAAAATTTTTTTTTTTTGGGGGGGGGGGCCCCCCCCC", "TTTTTTTTTTTGGGGGGGGGGGG")
#> Dead pointer detected. Attempting to reload...
#> Using tblastx Defaults...
#> Clock : ... seconds
#> 1
#> RecordBatchVector size: 1
#> Total rows across all batches: 1
#> [[1]]

# Expected empty results due to mismatched pointer type (Nucleotide vs Nucleotide with a blastp pointer) 
try(QuickBLAST::BLAST2Seqs(blastp_inst, "AAAAAAAAAAAATTTTTTTTTTTTGGGGGGGGGGGCCCCCCCCC", "TTTTTTTTTTTGGGGGGGGGGGG"))
#> [BLAST2Seqs()] No alignments could be computed
#> Clock : ... seconds
#> 1
#> RecordBatchVector size: 1
#> list()

# (~)Successful Protein-Protein search : Same protein
try(QuickBLAST::BLAST2Seqs(blastp_inst, "MQILLVEDDNTLFQELKKELEQWDFNVAGIEDFGKVMDTFESFNPEIVILDVQLPKYDGFYWCRKMREVSNVPILFLSSRDNPMDQVMSMELGADDYMQKPFYTNVLIAKLQAIYRRVYEFTAEEKRTLTWQDAVVDLSKDSIQKGDDTIFLSKTEMIILEILITKKNQIVSRDTIITALWDDEAFVSDNTLTVNVNRLRKKLSEISMDSAIETKVGKGYMAHE", "MQILLVEDDNTLFQELKKELEQWDFNVAGIEDFGKVMDTFESFNPEIVILDVQLPKYDGFYWCRKMREVSNVPILFLSSRDNPMDQVMSMELGADDYMQKPFYTNVLIAKLQAIYRRVYEFTAEEKRTLTWQDAVVDLSKDSIQKGDDTIFLSKTEMIILEILITKKNQIVSRDTIITALWDDEAFVSDNTLTVNVNRLRKKLSEISMDSAIETKVGKGYMAHE"))
#> Clock : ... seconds
#> 1
#> RecordBatchVector size: 1
#> Total rows across all batches: 1
#> [[1]]

# Successful NT-NT search
try(QuickBLAST::BLAST2Seqs(blastn_inst, "ATGGAGGAGCCGCAGTCAGATCCTAGCGTCGAGCCCCCTCTGAGTCAGGAAACATTTTCA", "ATGGAGGAGCCGCAGTCAGATCCTAGCCTCGAGCCCCCTCTGAGTCAGGAAACATTTTCA"))
#> Clock : ... seconds
#> 1
#> RecordBatchVector size: 1
#> Total rows across all batches: 1
#> [[1]]
#>  seq_info_num_alignments seq_info_seqids_qseqid seq_info_seqids_sseqid seq_info_seqs_qseq seq_info_seqs_sseq seq_info_strands seq_info_lengths_qlen
#> 1                       1                      3                      4                                                    +/+                    60
#>  seq_info_lengths_slen hsps_qhsp hsps_shsp hsps_pident hsps_pident_gap hsps_frames  hsps_evalue hsps_length hsps_length01 hsps_qstart hsps_qend
#>1                    60                        98.33333        98.33333         2/2 2.763249e-29          60             1           1        60
#>  hsps_sstart hsps_send hsps_bitscore hsps_score hsps_qcovhsp hsps_blast_score hsps_gaps hsps_nident hsps_mismatch hsps_positive hsps_n_splices
#>1           1        60      106.3793         57            0               57         0          59             1             0              0
#>  hsps_hsp_num hsps_sum_evalue hsps_product_coverage hsps_overall_identity hsps_negative_count hsps_matches hsps_high_quality_percent_coverage
#>1            1               0                     0                     0                   0            0                                  0
#>  hsps_exon_identity hsps_consensus_splices hsps_comp_adj_method
#>1                  0                      0                    0

```

## 3) Remote NCBI Searching

If you do not want to compile databases locally, QuickBLAST can interface directly with NCBI's remote servers.

```{r}
# Example of a remote timeout handling
try(QuickBLAST::RemoteBLAST(blastn_inst, query_input="AAAAAAAAAAAATTTTTTTTTTTTGGGGGGGGGGGCCCCCCCCC", database= "nr", input_type=1, return_values=TRUE))
#> Using blastn Defaults...
#> Max wait time: 360 seconds
#> 0%   10   20   30   40   50   60   70   80   90   100%
#> [----|----|----|----|----|----|----|----|----|----|
#> **************************************************|
#> RemoteBLAST(): C++ Runtime Error : Remote BLAST timed out
#> [1] FALSE

# Successful Remote Protein Query against the PDB database
try(QuickBLAST::RemoteBLAST(blastp_inst, query_input="MQILLVEDDNTLFQELKKELEQWDFNVAGIEDFGKVMDTFESFNPEIVILDVQLPKYDGFYWCRKMREVSNVPILFLSSRDNPMDQVMSMELGADDYMQKPFYTNVLIAKLQAIYRRVYEFTAEEKRTLTWQDAVVDLSKDSIQKGDDTIFLSKTEMIILEILITKKNQIVSRDTIITALWDDEAFVSDNTLTVNVNRLRKKLSEISMDSAIETKVGKGYMAHE", database= "pdb", input_type=1, return_values=TRUE))
#> Option: evalue set to : 100000
#> Max wait time: 360 seconds
#> 0%   10   20   30   40   50   60   70   80   90   100%
#> [----|----|----|----|----|----|----|----|----|----|
#> **************************************************|
#> (Success) Remote BLAST run ID: XDJNU28M014
#> Clock : ... seconds
#> RecordBatchVector size: 1
#> Total rows across all batches: 586
#> [[1]]
```

## 4) File-to-File Comparisons

For large-scale genomics, sequences are often stored in FASTA files. QuickBLAST can compare two files directly via multi-threaded reading, without loading everything into the R global environment first.

```{r}
# Create temporary FASTA files
query_fasta <- tempfile(fileext = ".fasta")
subject_fasta <- tempfile(fileext = ".fasta")

# Human TP53 gene fragment (Query)
writeLines(c(
  ">Human_p53_fragment\nATGGAGGAGCCGCAGTCAGATCCTAGCGTCGAGCCCCCTCTGAGTCAGGAAACATTTTCA"
), query_fasta)

# Slightly mutated version (Subject - e.g., an ortholog or variant)
writeLines(c(
  ">Variant_p53_fragment\nATGGAGGAGCCGCAGTCAGATCCTAGCCTCGAGCCCCCTCTGAGTCAGGAAACATTTTCA"
), subject_fasta)

# Efficiently compare the files using the BLASTN instance
file_results <- QuickBLAST::BLAST2Files(
  blastn_inst, 
  query = query_fasta, 
  subject = subject_fasta,
  return_values = TRUE,
  num_threads = 1
)
#>Num Threads: 1
#>Total Records (Q + S): 2 (1 + 1)
#>Batch Size: 2
#>Computing: [========================================] 100% (done)                         
#>Total Records Processed: 2
#>Clock : ... seconds
#>RecordBatchVector size: 1
#>Total rows across all batches: 1

print(file_results)
#>[[1]]
#>  seq_info_num_alignments seq_info_seqids_qseqid seq_info_seqids_sseqid seq_info_seqs_qseq
#>1                       1     Human_p53_fragment   Variant_p53_fragment                   
#>  seq_info_seqs_sseq seq_info_strands seq_info_lengths_qlen seq_info_lengths_slen
#>1                                 +/+                    60                    60
#>                                                     hsps_qhsp
#>1 ATGGAGGAGCCGCAGTCAGATCCTAGCGTCGAGCCCCCTCTGAGTCAGGAAACATTTTCA
#>                                                     hsps_shsp hsps_pident hsps_pident_gap
#>1 ATGGAGGAGCCGCAGTCAGATCCTAGCCTCGAGCCCCCTCTGAGTCAGGAAACATTTTCA    98.33333        98.33333
#>  hsps_frames  hsps_evalue hsps_length hsps_length01 hsps_qstart hsps_qend hsps_sstart
#>1         2/2 2.763249e-29          60             1           1        60           1
#>  hsps_send hsps_bitscore hsps_score hsps_qcovhsp hsps_blast_score hsps_gaps hsps_nident
#>1        60      106.3793         57            0               57         0          59
#>  hsps_mismatch hsps_positive hsps_n_splices hsps_hsp_num hsps_sum_evalue
#>1             1             0              0            1               0
#>  hsps_product_coverage hsps_overall_identity hsps_negative_count hsps_matches
#>1                     0                     0                   0            0
#>  hsps_high_quality_percent_coverage hsps_exon_identity hsps_consensus_splices
#>1                                  0                  0                      0
#>  hsps_comp_adj_method
#>1                    0
```

## 5) Local Database Creation and Searching

For high-throughput homology searches against a fixed reference, compiling the sequences into an indexed BLAST database is much faster than raw file-to-file comparisons. QuickBLAST wraps the NCBI makeblastdb functionality directly into R.


```{r}
db_path <- tempfile()

# 1. Compile the local BLAST database
# db_type can be "nucl" or "prot"
QuickBLAST::MakeBLASTDB(
  ptr = blastn_inst,
  input_file = subject_fasta,
  database_name = file.path(tempdir(),"My_Test_DB"),
)

#>Building a new DB, current time: ...
#>New DB name:   /tmp/Rtmp8UacQU/My_Test_DB
#>New DB title:  /tmp/RtmpjpBk7u/file214ace524c2945.fasta
#>Sequence type: Nucleotide
#>Keep MBits: T
#>Maximum file size: 3000000000B
#>Adding sequences from FASTA; added 1 sequences in ... seconds.
#>
#>
#>makeblastdb finished successfully
#>[1] "/tmp/Rtmp8UacQU/My_Test_DB"
 
# 2. Search a query file against the custom local database
db_results <- QuickBLAST::BLAST2DBs(
 blastn_inst, 
 query = query_fasta, 
 subject = file.path(tempdir(),"My_Test_DB"),
 return_values = TRUE,
 num_threads = 2
)
 
#>Building a new DB, current time: ...
#>New DB name:   /tmp/Rtmp8UacQU/file22d37f699b82f7.fasta.db
#>New DB title:  /tmp/Rtmp8UacQU/file22d37f699b82f7.fasta
#>Sequence type: Nucleotide
#>Keep MBits: T
#>Maximum file size: 3000000000B
#>Adding sequences from FASTA; added 1 sequences in ... seconds.
#>
#>
#>makeblastdb finished successfully
#>Q :/tmp/Rtmp8UacQU/file22d37f699b82f7.fasta.db
#>S :/tmp/Rtmp8UacQU/My_Test_DB
#>Num Threads: 2
#>Total Records (Q + S): 2 (1 + 1)
#>Batch Size: 3
#>0%   10   20   30   40   50   60   70   80   90   100%
#>[----|----|----|----|----|----|----|----|----|----|
#>Batch Hits: 1
#>**************************************************|
#>Processed Batches:1
#>Total Records Processed: 1
#>Clock : ... seconds
#>RecordBatchVector size: 1
#>Total rows across all batches: 1
 
print(db_results)
#>[[1]]
#>  seq_info_num_alignments seq_info_seqids_qseqid seq_info_seqids_sseqid
#>1                       1     Human_p53_fragment     Human_p53_fragment
#>  seq_info_seqs_qseq seq_info_seqs_sseq seq_info_strands seq_info_lengths_qlen
#>1                                                    +/+                    60
#>  seq_info_lengths_slen                                                    hsps_qhsp
#>1                    60 ATGGAGGAGCCGCAGTCAGATCCTAGCGTCGAGCCCCCTCTGAGTCAGGAAACATTTTCA
#>                                                     hsps_shsp hsps_pident
#>1 ATGGAGGAGCCGCAGTCAGATCCTAGCGTCGAGCCCCCTCTGAGTCAGGAAACATTTTCA    98.33333
#>  hsps_pident_gap hsps_frames  hsps_evalue hsps_length hsps_length01 hsps_qstart
#>1        98.33333         1/1 2.763249e-29          60             1           0
#>  hsps_qend hsps_sstart hsps_send hsps_bitscore hsps_score hsps_qcovhsp
#>1        59           0        59      106.3793         57            0
#>  hsps_blast_score hsps_gaps hsps_nident hsps_mismatch hsps_positive hsps_n_splices
#>1               57         0          59             1             0              0
#>  hsps_hsp_num hsps_sum_evalue hsps_product_coverage hsps_overall_identity
#>1            1               0                     0                     0
#>  hsps_negative_count hsps_matches hsps_high_quality_percent_coverage
#>1                   0            0                                  0
#>  hsps_exon_identity hsps_consensus_splices hsps_comp_adj_method
#>1                  0                      0                    0


```

## 6) High-Performance Output with Apache Arrow

By default, the return_values = TRUE flag seen above brings data directly into R as an Rcpp::List. However, when processing millions of alignments, memory can max out quickly.

To solve this, QuickBLAST writes results asynchronously using Apache Arrow formats (arrow::ipc or arrow::parquet). This creates a highly compressed, blazingly fast columnar storage file instead of a clunky TSV or CSV.


```{r}
# Specify an output path
out_parquet <- tempfile(fileext = ".parquet")

# Perform the search and write straight to an Arrow IPC file
QuickBLAST::BLAST2Files(
  blastn_inst, 
  query = query_fasta, 
  subject = subject_fasta,
  out_file = out_parquet,
  out_format = "parquet",
  return_values = FALSE # We don't need it in RAM, we are writing to disk
)
#>Writing to : /tmp/Rtmp8UacQU/file22d37f7207ae58.parquet
#>Output Format : parquet
#>Num Threads: 2
#>Total Records (Q + S): 2 (1 + 1)
#>Batch Size: 3
#>Computing: [========================================] 100% (done)                         
#>/tmp/Rtmp8UacQU/file22d37f7207ae58.parquet
#>Done writing to file.
#>Total Records Processed: 2
#>Clock : ... seconds
#>[1] TRUE

# Load the hits back into an R data.frame natively via QuickBLAST
hits_df <- QuickBLAST::LoadBLASTHits(out_parquet, format = "parquet")
head(hits_df)
#># A tibble: 1 × 2
#>  seq_info$num_alignments $seqids$qseqid     $seqs$qseq $strands $lengths$qlen hsps$qhsp
#>                    <int> <chr>              <chr>      <chr>            <int> <chr>    
#>1                       1 Human_p53_fragment ""         +/+                 60 ""       
#># ℹ 33 more variables: seq_info$seqids$sseqid <chr>, seq_info$seqs$sseq <chr>,
#>#   seq_info$lengths$slen <int>, hsps$shsp <chr>, $pident <dbl>, $pident_gap <dbl>,
#>#   $frames <chr>, $evalue <dbl>, $length <int>, $length01 <dbl>, $qstart <int>,
#>#   $qend <int>, $sstart <int>, $send <int>, $bitscore <dbl>, $score <dbl>, $qcovhsp <dbl>,
#>#   $blast_score <dbl>, $gaps <int>, $nident <int>, $mismatch <int>, $positive <int>,
#>#   $n_splices <int>, $hsp_num <int>, $sum_evalue <dbl>, $product_coverage <dbl>,
#>#   $overall_identity <dbl>, $negative_count <int>, $matches <dbl>, …
```

## Conclusion

QuickBLAST modernizes sequence alignment in R. By managing your instances appropriately, keeping data in native C++ structures, and offloading massive search results to Apache Arrow formats, you can run high-throughput bioinformatic pipelines natively without relying on slow Sys.Call() bash wrappers.
